---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
setwd("/Users/Fahimeh/work/Data-Science/Coursera")
load("gss.Rdata")
```



* * *

## Part 1: Data
We can assume that the response of each person is independent of another person. 

* * *

## Part 2: Research question
Q1. Child of educated(after 12 grade) mothers are more likely to go to college?

## Part 3: Exploratory data analysis

```{r}
# Lets define a new variable to call mother and child educated or not educated. 
# We will consider a mother and a child educated if they go to college or higher
gss1 <- gss %>%
  mutate(childedu = ifelse(degree == "Junior College" | degree == "Bachelor" | degree == "Graduate", "educated_child", "noteducated_child")) %>% 
  mutate(macol = ifelse(maeduc > 12, "educated_mother","noteducated_mother")) %>%
  mutate(pacol = ifelse(paeduc > 12, "educated_father","noteducated_father"))
  
educ_childof_educ_ma <- gss1 %>% 
  filter(childedu == "educated_child", macol == "educated_mother") %>%
  summarise(n())

total_childof_educ_ma <- gss1 %>% 
  filter(macol == "educated_mother") %>%
  summarise(n())

noteduc_childof_educ_ma <- gss1 %>% 
  filter(childedu == "noteducated_child", macol == "educated_mother") %>%
  summarise(n())

total_childof_noteduc_ma <- gss1 %>% 
  filter(macol == "noteducated_mother") %>%
  summarise(n())

P_hat_educ_childof_educ_ma <- educ_childof_educ_ma / (educ_childof_educ_ma + noteduc_childof_educ_ma)

educ_childof_noteduc_ma <- gss1 %>% 
  filter(childedu == "educated_child", macol == "noteducated_mother") %>%
  summarise(n())

noteduc_childof_noteduc_ma <- gss1 %>% 
  filter(childedu == "noteducated_child", macol == "noteducated_mother") %>%
  summarise(n())

P_hat_educ_childof_noteduc_ma <- educ_childof_noteduc_ma / (educ_childof_noteduc_ma + noteduc_childof_noteduc_ma)

# Proportion of educated child when mother is educated
P_hat_educ_childof_educ_ma

#Proportion of educated child when mother is not educated
P_hat_educ_childof_noteduc_ma

# Difference between these two proportion
P_hat_educ_childof_educ_ma - P_hat_educ_childof_noteduc_ma
```

## Part 4: Inference

There is a distinct difference in the proportion of educated when the mother is educated or not educated (We consider degrees after high school as a educated degree). But is it statistically significant? Lets do a hypthesis test:

```{r}
# H0 : P_hat_educ_childof_educ_ma = P_hat_educ_childof_noteduc_ma
# HA : P_hat_educ_childof_educ_ma > P_hat_educ_childof_noteduc_ma
# First lest compute the standard error 
#SE = sqrt(p_pool * (1-p_pool)/ n1 + p_pool * (1-p_pool)/ n2)

p_pool <- (educ_childof_educ_ma + educ_childof_noteduc_ma) / (total_childof_educ_ma + total_childof_noteduc_ma)
SE <- sqrt( p_pool * (1 - p_pool) / total_childof_educ_ma + p_pool * (1 - p_pool) / total_childof_noteduc_ma) 

(P_hat_educ_childof_educ_ma - P_hat_educ_childof_noteduc_ma) / SE
pnorm (57.97559, lower.tail = FALSE)

# An alternative way of doing it in R
inference(childedu, macol, gss1, type = "ht", statistic = "proportion", success = "educated_child", method = "theoretical", null = 0, alternative = "greater")
 
# Since p-value is small we reject the null hypothesis and we accept the alternative hypothesis 
```
We are 95 percent confident that the propotion of educated child from educated mothers are higher than not educated mothers. We can also define a confidence interval for the proportion difference:

```{r}
diff <- P_hat_educ_childof_educ_ma - P_hat_educ_childof_noteduc_ma
c(diff + qnorm(0.975) * SE , diff - qnorm(0.975) * SE)

# Or using R:
inference(childedu, macol, gss1, type = "ci", statistic = "proportion", success = "educated_child", method = "theoretical")
```

We are 95% confidence that the proportion of educated child coming from educated mother is between 28 to 30 percent higher than educated child coming from not educated mother.

Q2. Are children of educated father more likely to go to college?

```{r}
educ_childof_educ_pa <- gss1 %>% 
  filter(childedu == "educated_child", pacol == "educated_father") %>%
  summarise(n())

total_childof_educ_pa <- gss1 %>% 
  filter(pacol == "educated_father") %>%
  summarise(n())

noteduc_childof_educ_pa <- gss1 %>% 
  filter(childedu == "noteducated_child", pacol == "educated_father") %>%
  summarise(n())

total_childof_noteduc_pa <- gss1 %>% 
  filter(pacol == "noteducated_father") %>%
  summarise(n())

P_hat_educ_childof_educ_pa <- educ_childof_educ_pa / (educ_childof_educ_pa + noteduc_childof_educ_pa)

educ_childof_noteduc_pa <- gss1 %>% 
  filter(childedu == "educated_child", pacol == "noteducated_father") %>%
  summarise(n())

noteduc_childof_noteduc_pa <- gss1 %>% 
  filter(childedu == "noteducated_child", pacol == "noteducated_father") %>%
  summarise(n())

P_hat_educ_childof_noteduc_pa <- educ_childof_noteduc_pa / (educ_childof_noteduc_pa + noteduc_childof_noteduc_pa)

# Proportion of educated child when mother is educated
P_hat_educ_childof_educ_pa

#Proportion of educated child when mother is not educated
P_hat_educ_childof_noteduc_pa

# Difference between these two proportion
P_hat_educ_childof_educ_pa - P_hat_educ_childof_noteduc_pa

inference(childedu, pacol, gss1, type = "ht", statistic = "proportion", success = "educated_child", method = "theoretical", null = 0, alternative = "greater")

inference(childedu, pacol, gss1, type = "ci", statistic = "proportion", success = "educated_child", method = "theoretical")
```

With 95% confidence, the propotion of educated children of educated fathers are 0.31 to 0.34 more than the propotion of not educated fathers.


